Is malaria control profitable? Return on investment of privately-managed residential fumigations at a large sugarcane processing facility in Southern Mozambique

### 2018-04-20 inst/rmd/02_methods.Rmd
changes

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed interrupted time series [@Lopez_Bernal_2016] and a linear probability approach using the following econometric model.

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{t}) + (\beta_2{IRS}*\beta_3{IRS_t}) + ... + \epsilon
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention was not a simple yes/no, but rather the product of whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

@@

All data processing and analysis were carried out in R [@R] and all analysis code is freely available online [@brewgit]. Ethical approval for this project was obtained from the Institutional Ethics Review Board for Health at the CISM prior to data collection.

## Descriptive statistics

Age, types of workers, bla bla bla. \todo{Will add a table here and some overview}

>>>>

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed linear mixed effects models [@Minalu2011] [@Liu2010] (with individual worker fixed effects, and random effects for other covariates)  in lieu of interrupted time series [@Lopez_Bernal_2016] so as to explicitly adjust for confounders [@Bell2014]. We divide into 4 different models for different worker types, so as to account for the potential time-varying effects of worker type on risk of malaria. Our model specification is as follows

$$
\hat{Y_{it}} = \hat{\beta}_{0} +  \hat{\beta}_{1}\text{Season}_{t} * (\hat{\beta}_2{IRS_{it}}*\hat{\beta}_3{IRStime_{it}}) + \alpha_i + \delta_t + \upsilon_{it}
$$

$\hat{Y}_{it}$ is the rate of absence. $\beta_{1}$ is the binary "season" variable, imputed from overall district clinical incidence. Our intervention is whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment, represented, respectively, by $\beta_{2}$ and $\beta_{3}$. $\alpha_i$ reprsents the time invariant worker fixed effects, and $\delta_i$ represents the fixed effect of the particular malaria season. $\upsilon$ is the error term.

We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation; though its lagged effect on malaria incidence is likely captured by the seasonality term, we include rainfall since it's immediate effect (through its impact on transportation and working conditions) may also affect absenteeism.

Details will go here about externalities (neighborhood effects, positive spillover to district, negative spillover from district, how large firm may mean that most positive externalities are absorbed by the firm, and not many negative externalities, but maybe with a smaller firm it wouldn't work as well)\todo{NOT DONE YET.}

@@

All data processing and analysis were carried out in R [@R] and all analysis code is freely available online [@brewgit]. Ethical approval for this project was obtained from the Institutional Ethics Review Board for Health at the CISM (CIBS-CISM) prior to data collection.

\newpage

### 2018-04-20 inst/rmd/03_results.Rmd
changes

@@

**Effect of IRS on absenteeism**: Immediately following IRS, a worker's likelihood of absence drops significantly (figure x, panel i). As one would expect if the mechanism by which IRS reduces absence is through reduced malaria infection, the effect of IRS during the low transmission season is significant, but far less substantial in effect size (figure x, panel ii).

```{r, fig.height = 3.5, fig.width = 6, fig.cap = 'i. Absenteeism before and after IRS administration for all workers who ever received IRS; ii. The same, but segregated by malaria and non-malaria seasons'}
plot_data <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs, c("date", "unidade")) %>%
  mutate(months_since = days_since %/% 30) %>%
  filter(!is.na(months_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick))

x <- plot_data %>%
  group_by(months_since) %>%
  summarise(absences = length(which(absent)),
            sick_absences = length(which(absent_sick)),
            eligibles = length(absent)) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) %>%
  mutate(sick_absenteeism_rate = sick_absences / eligibles * 100)

g1 <- ggplot(data = x,
       aes(x = months_since,
           y = absenteeism_rate)) +
  geom_step(color = 'darkred', alpha = 0.8) +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) +
  labs(x = 'Months relative to IRS',
       y = 'Absenteeism rate') +
  theme_maragra()

plot_data2 <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs, by = c("date", "unidade")) %>%
  mutate(months_since = days_since %/% 30) %>%
  filter(!is.na(months_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick))

x2 <- plot_data2 %>%
  group_by(months_since, time_period) %>%
  summarise(absences = length(which(absent)),
            sick_absences = length(which(absent_sick)),
            eligibles = length(absent)) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) %>%
  mutate(sick_absenteeism_rate = sick_absences / eligibles * 100) %>%
  mutate(before = months_since < 0)

g2 <- ggplot(data = x2 %>% mutate(time_period = paste0(time_period)),
       aes(x = months_since,
           y = absenteeism_rate)) +
  geom_step(color = 'darkred', alpha = 0.8) +
  facet_wrap(~time_period, ncol = 1) +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) +
  labs(x = 'Months relative to IRS',
       y = 'Absenteeism rate') +
  theme_maragra()
Rmisc::multiplot(g1, g2, cols = 2)
```

@@

The model results for sickness only absenteeism (table 5) suggest that IRS has a significant effect on permanent field workers. The effect is directionally similar among temporary fieldworkers (but does not reach the level of significance). Among non-fieldworkers, the effect of IRS is mixed.

@@

## Externalities

This section is in progress. Looking at neighborhood effects, core-periphery effects, etc.

>>>>

## Descriptive statistics

```{r}
# Calculate number of workers
n <- length(unique(ab_panel$oracle_number[ab_panel$date >= '2012-01-01' & ab_panel$date <= '2016-12-31']))
```

We collected absenteeism and demographic data data on `r n` workers from 2012 through 2016. Workers were approximately 60% male, and more than 80% fieldworkers (predominantly cane-cutters). Most were in their 20s and 30s and employed on temporary contracts. Table 1 shows overall worker details divided by "Treatment" versus "Control" status. "Treatment" is considered the time beginning at IRS administration and ending one year later; "Control" is considered time observed either prior to IRS administration (and greater than 1 year after previous IRS administration) or among workers who never received IRS. Because of the temporal nature of these categories, many workers belong to both treatment and control groups, albeit at different times.

```{r, results='asis'}
# x <- ab_panel %>%
#   left_join(workers %>%
#               dplyr::select(oracle_number, date_of_birth, permanent_or_temporary, sex))
x <- model_data
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = -1)
x$status <- x$permanent_or_temporary

calculate <- function(var= 'sex',
                      data){
  out <- 
    data %>%
    group_by_(var) %>%
    summarise(`Treatment workers` = length(unique(oracle_number[!days_since %in% c('Never', 'Before')])),
              `Treatment days` = length(oracle_number[!days_since %in% c('Never', 'Before')]),
              `Control workers` = length(unique(oracle_number[days_since %in% c('Never', 'Before')])),
              `Control days` = length(oracle_number[days_since %in% c('Never', 'Before')])) %>%
    ungroup() %>%
    # mutate(p = round(n / sum(n) * 100, digits = 2)) %>%
    mutate(variable = '') 
    names(out)[1] <- ' '
    out$variable[1] <- Hmisc::capitalize(var)

    out <- out %>% dplyr::select(variable, ` `, `Treatment workers`, `Treatment days`, `Control workers`, `Control days`)
    out$` ` <- as.character(out$` `)
  return(out)
}
out_list <- list()
vars <- c('department', 
           'sex',
           'status',
           'age')
for(i in 1:length(vars)){
  message(i)
  y <- calculate(var = vars[i], data = x)
  out_list[[i]] <- y
}
demo_table <- bind_rows(out_list)
names(demo_table)[1:2] <- c('Variable', ' ')
print(xtable::xtable(demo_table,
                     caption = 'Overall worker characteristics',
                     align = c('r', 'r', rep('l', 5))),
      include.rownames = FALSE)

```

Temporary workers tended to be younger and male; female temporary workers being on average 5-10 years older than their male counterparts. Permanent workers had a bi-modal age distribution, the older peak explained by the greater density of males in administrative roles. Females accounted for 43% of temporary workers but only 16% of permanent workers (figure x).

```{r, fig.width = 5, fig.height = 3.5, fig.align='center', fig.cap = 'Age distribution by sex and worker status'}
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = 2)
g1 <- ggplot(data = x,
       aes(x = age,
           group = sex,
           fill = sex)) +
  geom_density(alpha = 0.8, n = 2^4) +
  facet_wrap(~permanent_or_temporary) +
  theme_maragra() +
  labs(x = 'Age',
       y = 'Density') +
  scale_fill_manual(name = 'Sex',
                    values = c('lightblue', 'grey'))
g1
```

During the study period, weather followed typical seasonal trends for the area, albeit slightly drier than previous periods.

```{r, fig.width = 6, fig.height = 3.5, fig.align='center', fig.cap = 'i. Monthly total precipitation in the Maniça district; ii. Average monthly temperature (bars) during the study period, as well as monthly maximum (red) and minimum (blue) temperatures'}
x <- weather %>%
  mutate(month = format(date, '%m')) %>%
  mutate(year = format(date, '%Y')) %>%
  mutate(date = as.Date(paste0(year, '-', month, '-01'))) %>%
  group_by(date) %>%
  summarise(precipitation = sum(precipitation, na.rm = TRUE),
            temp_max = max(temp_max, na.rm = TRUE),
            temp_min = min(temp_min, na.rm = TRUE),
            temp = mean(temp))
# x <- gather(x, key, value,precipitation:temp)

g1 <- ggplot(data = x,
       aes(x = date,
           y = precipitation)) +
  geom_bar(stat = 'identity',
           fill = 'lightblue',
           size = 0.1) +
  theme_maragra() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = 'Quarter',
       y = 'Milimeters') +
  scale_x_date(labels = scales::date_format("%Y"))

g2 <- ggplot(data = x,
            aes(x = date,
                y = temp)) +
  geom_bar(stat = 'identity', fill = 'grey', size = 0.1) +
  geom_line(aes(y = temp_max),
             color = 'darkred', 
             alpha = 0.5) +
  geom_line(aes(y = temp_min),
             color = 'blue', 
             alpha = 0.5) +
  # geom_linerange(aes(ymin = temp_min,
  #                    ymax = temp_max),
  #                alpha = 0.6,
  #                color = 'blue') +
   theme_maragra() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = 'Quarter',
       y = 'Degrees (Celcius)') +
  scale_x_date(labels = scales::date_format("%Y"))
Rmisc::multiplot(g1, g2, cols = 2)
```

@@

## Analysis

### Effect of IRS on absenteeism

Immediately following IRS, a worker's likelihood of absence drops significantly (appendix figure, panel A). As one would expect if the mechanism by which IRS reduces absence is through reduced malaria infection, the effect of IRS during the low transmission season is significant, but far less substantial in effect size (appendix figure, panel B).

@@

The model results for sickness only absenteeism (table 5) suggest that IRS has a significant effect on permanent field workers. The effect is directionally similar among temporary fieldworkers (but does not reach the level of significance). Among non-fieldworkers, the effect of IRS is mixed.\todo{I'll probably remove this from the final version, since we're saying that the data aren't that good...}

@@

### Externalities

One potential shortcoming of our analysis thus far is that a "control" worker (ie, one whose house has never been sprayed or was sprayed greater than one year ago) may actually be conferred some protection against malaria by the fumigations at his/her neighbors' houses. This positive spillover effect would theoretically go through two channels: (i) via a reduction of mosquitoes in the vicinity and (ii) via a reduction of the malaria parasite in the blood of humans in the vicinity (ie, the parasite "reservoir").

By the same token, just as being surrounded by IRS-treated homes may indirectly protect someone living in a home not treated by IRS, one might surmise that living on the periphery of the company grounds might lead to more malaria risk relative to living in the core, since homes near the periphery but outside of Maragra are less likely to be fumigated. In other words, to the extent that both positive and spillover effects are significant enough to warrant inclusion in our model, we would expect that (i) the prevalence of IRS in a worker's neighborhood should be associated with a reduction in absenteeism of the worker and (b) workers living further from the periphery of the company grounds should have lower absenteeism than workers living at the periphery. We test both of these hypotheses.

#### Neighborhood protective effect

\todo{Working on this now}

#### Core-periphery difference

\todo{Working on this now}

### 2018-04-20 inst/rmd/paper.Rmd
changes

```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE, cache = FALSE}
# Packages
library(tidyverse)
library(knitr)
library(Hmisc)
library(brew)
library(maragra)
library(knitr)
library(RColorBrewer)
library(extrafont)
library(kableExtra)
loadfonts()
## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev = "cairo_pdf",
               fig.pos="h",
               fig.path = 'figures/')
opts_knit$set(width=75)
options(xtable.comment = FALSE)

>>>>

```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE, cache = FALSE}
# Packages
library(tidyverse)
library(knitr)
library(Hmisc)
library(brew)
library(maragra)
library(knitr)
library(RColorBrewer)
library(extrafont)
library(kableExtra)
loadfonts()
## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev = "cairo_pdf",
               fig.pos="!h",
               fig.path = 'figures/')
opts_knit$set(width=75)
options(xtable.comment = FALSE)
